---
title: "esm206_lab9- Part 2 Chi-square with Lizards"
author: "Abigail Sanford"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Attach packages
library(here)
library(tidyverse)
library(janitor)
library(broom)
```

## Read in the data

```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```


# Find counts of lizards in the dataset by species

```{r}
lizards %>% 
  count(spp) %>% 
  arrange(-n) # arrange in decreasing order by variable `n` that contains the count. Now you can see the count of lizards studied for each species
```


## Data wrangling: Make a subset

- Only want a subset with UTST lizards (sideblotched lizards)
- Only keep sites with 3 highest trap populations: IBPE (enclosure grasslands site), RABB (rabbit mesquite site), "EAST" (east tarbush site).
- Keep observations where the tail is recorded. (either W = whole, or B = broken)

```{r}
utst <- lizards %>% 
  filter(spp == "UTST") %>% # filter to only keep observation where the species column matches UTST
  filter(site %in% c("IBPE", "RABB", "EAST")) %>% # filter to only keep sites IBPE, RABB, and EAST. Use the in operator to say 'look for any match within the vector that follows'
  filter(tail %in% c("B", "W")) # only keep broken or whole tail observations, exclude NA values.
```


Reminder: if you want to check the different unique levels that exists within a variable we can go to the console and type: unique(utst$site)
 or unique(data_set$column_name)
 
 
 ## Find Counts and Proportions of Lizards with Broken or Whole Tails and Each Site 
 
 These are all sideblotched lizards, so just a reminder to keep that in mind when doing a write up.
 
1) normal way: use count function from dplyr and say I want to count by site and tail columns. which will tell me, for each site how many lizards had a recorded broken tail and how many lizards had a recorded whole tail.

- It comes out difficult to read and kind of ugly. It's good for data analysis (cause it's very obvious and long what each measurement is coming from), but is bad for communication. It has the right information, but there's a better way that we can do this.

sidenote: you can use pivotwinder to change the tail column into two different outcomes, and then have the counts listed that way
 
```{r}
utst %>% 
  count(site,tail) # count observations in site and tail

```
 
 
2) use janitor that returns counts in contingency table format. WE can also use it to show proportion directly, which is really useful when we are going to be using a chi-squared test.

- Rstudio's default is to calculate precentages (proportions) by rows. We can change that.

```{r}
tail_count <- utst %>% 
  janitor::tabyl(site, tail) # create contingency table dataframe using the janitor package and tabyl function. we put the `janitor::` to tell rstudio which package we want it to use.

tail_proportions <- tail_count %>% 
  janitor::adorn_percentages() %>% # shows percentage of lizards at each site
  janitor::adorn_pct_formatting(digits = 2) %>% # round by 2 decimal places
  janitor::adorn_ns() # get the actual count for each tail

tail_proportions # then I could make a nice pretty table using all the other tricks we have used previously to make nice looking tables

```

 
Now we want to test: Do we have enough evidence to reject the null hypothesis that tail breakage is independent of site?

first we need to convert the information in tail_count data frame into a true contingency table. Which means only the count information exists in the table. In the current tail_count data frame, the sites column is being read as data. It is site information that is additional information to the counts. Which we don't want. So we are going to use a function called `column_to_rowname` that converts the column we want into row names instead so all we are left with is the contingency table with counts.

## Convert to a contingency table

Use `column_to_rowname()` function to do that. 

```{r}
lizard_ct <- tail_count %>% 
  column_to_rownames(var = "site") # variable = "column_name_to_convert". Quotations indicate we are removing it as a variable, and we are using the inputs for row names instead. 
```

Once we have our data in the format where there is only count data, we can run a chisquared test. 

## Chi-square

Are tail breakage and site independent. If the sites are independent, we would expect very little difference in the count proportions between the three sites. 

Chi square can be challenging because we can ask the same question in many different ways. But once we have it in the right format, and know what question to ask, the code is straightforward.

```{r}
lizard_x2 <- chisq.test(lizard_ct)

lizard_x2 # gives you the chisquared test statistic value, the aggregrate value how different the values in our tables are from the 'perfect situation' values.degrees of freedom = ((number_of_rows - 1) * (number_of_columns - 1)). p-value, a probability of 0.0014

lizard_tidy <- broom::tidy(lizard_x2) # creates a new dataframe that is a tidy dataframe that we can easily pull info from for inline referencing. 

```


Use a significance level of .05, yes we have sufficient evidence to reject the null hypothesis of evidence. SO we retain the alternative hypothesis that site and tail breakage are not independent.
 
Different ways to write it: 

There is a significant association (i.e. non-independence) between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`) = `r round(lizard_tidy$statistic,2)`, *p* = `r round(lizard_tidy$p.value, 3)`).

lizard_tidy$parameter: gives you the degrees of freedom
lizard_tidy$statistic: 
*p* = `r lizard_tidy$p.value: gives you the p-value

- Site and lizard tail breakage are not independent.

 
 









